Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-31 articles/cycle/alps-cycle.tex
Make last minor changes before resubmission.

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
    The EPICA simulations yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a retreat and then a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, red curves) in a very good agreement with
    these geological data. On the other hand, the simulations forced by the
    GRIP palaeo-temperature record yield two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat,
    in contrast to dated records of large-scale ice retreat by 17 to
    16\,ka\citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially sculpted
    valley troughs, represent the maximum elevation of the LGM ice surface,
    maximum ice thickness in the Alps has been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geomorphological and cosmogenic
    nuclide dating evidence from
    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994}, the
    British Isles \citep[e.g.,][]{Fabel.etal.2012, Ballantyne.Stone.2015}, and
    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved
    under cold-based ice, sometimes for several glacial cycles
    \citep{Stroeven.etal.2002}. Thus, the trimline could also mark the
    maximum elevation of the transition from temperate to cold-based ice, or a
    late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
    However, no such evidence has been reported in the Alps.

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.
    Interestingly, tree and fire remains evidence nunataks in the south-western
    Alps by 21\,ka \citep{Carcaillet.Blarquez.2017, Carcaillet.etal.2018}.
    However, regional sensitivity tests show that a modelled surface compatible
    with the trimline elevations is incompatible with the documented
    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2017}.

@@

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
        Tagliamento Glaciers.
        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
        time, bilinearly interpolated along the corresponding profiles, showing
        numerous cycles of advance and retreat over the last glacial cycle
        modulated by subglacial topography and catchment geometry. Shaded gray
        areas indicate the timing for MIS~2 and MIS~4
        \citep{Lisiecki.Raymo.2005}. Isolated patches indicate episodic
        advances from tributary glaciers.}
      \label{fig:profiles}
    \end{figure*}

@@

    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
    Aschwanden for providing constant help and development with PISM, in
    particular with recent issues with the computation of ice temperature
    (Github issue no.~371) and with the computation of bedrock deformation
    (Github issues no.~370 and 377). We are equally thankful to three
    anonymous reviewers, Giovanni Monagato, and the editor Andreas Vieli for
    their careful read and constructive inputs during peer-review.
    We thank Marc Luetscher for an insightful discussion of
    preliminary results during the EGU~General Assembly~2017. The experimental
    design and article layout used here are based on previous work which
    received much guidance from Irina Rogozhina and Arjen~P. Stroeven.
    The current work was supported by the Swiss National Science Foundation
    grants no.~200020-169558 and 200021-153179/1 to M.~Funk. Computer resources
    were provided by the Swiss National Supercomputing Centre (CSCS)
    allocations no.~s573 and sm13 to J.~Seguinot.

>>>>

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
    The EPICA simulations yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a retreat and then a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, red curves) in a very good agreement with
    these geological data. On the other hand, the simulations forced by the
    GRIP palaeo-temperature record yield two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat,
    in contrast to dated records of large-scale ice retreat by 17 to
    16\,ka \citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially sculpted
    valley troughs, represent the maximum elevation of the LGM ice surface,
    maximum ice thickness in the Alps has been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geomorphological and cosmogenic
    nuclide dating evidence from
    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994}, the
    British Isles \citep[e.g.,][]{Fabel.etal.2012, Ballantyne.Stone.2015}, and
    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved
    under cold-based ice, sometimes for several glacial cycles
    \citep{Stroeven.etal.2002}. Thus, the trimline could also mark the
    maximum elevation of the transition from temperate to cold-based ice, or a
    late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
    However, no such evidence has been reported in the Alps. Instead, tree and
    fire remains record the presence of nunataks in the south-western Alps by
    21\,ka \citep{Carcaillet.Blarquez.2017, Carcaillet.etal.2018}.

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here.
    However, regional model sensitivity tests show that a modelled surface
    compatible with the trimline elevations is incompatible with the documented
    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
    is not doable in the Alps due its lower values, active tectonics and
    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.

@@

    \begin{figure*}[p]
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
        Tagliamento Glaciers.
        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
        time, bilinearly interpolated along the corresponding profiles, showing
        numerous cycles of advance and retreat over the last glacial cycle
        modulated by subglacial topography and catchment geometry. Shaded gray
        areas indicate the timing for MIS~2 and MIS~4
        \citep{Lisiecki.Raymo.2005}. Isolated patches indicate episodic
        advances from tributary glaciers.}
      \label{fig:profiles}
    \end{figure*}

@@

    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
    Aschwanden for providing constant help and development with PISM, in
    particular with recent issues with the computation of ice temperature
    (Github issue no.~371) and with the computation of bedrock deformation
    (Github issues no.~370 and 377). We are equally thankful to three
    anonymous reviewers, Giovanni Monagato, and the editor Andreas Vieli for
    their meticulous readings and constructive inputs during peer-review.
    We thank Marc Luetscher for an insightful discussion of
    preliminary results during the EGU~General Assembly~2017. The experimental
    design and article layout used here are based on previous work which
    received much guidance from Irina Rogozhina and Arjen~P. Stroeven.
    The current work was supported by the Swiss National Science Foundation
    grants no.~200020-169558 and 200021-153179/1 to M.~Funk. Computer resources
    were provided by the Swiss National Supercomputing Centre (CSCS)
    allocations no.~s573 and sm13 to J.~Seguinot.
